---
title: "Untitled"
author: "Reuben"
date: "2024-12-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Clear environment
```{r}
rm(list = ls())
```

```{r}
# Read in data
rna <- read.csv("", row.names = 1)# add directory to rna data - data used here is from S-CORT
# Check for duplicate genes
sum(duplicated(rownames(rna)))
# Check for duplicate samples
sum(duplicated(colnames(rna)))
# Check for na
sum(is.na(rna))

```
# Remove sex chromosome genes
```{r}
# Read the txt file of gene names found on the x and y chromosomes
sex_chrom_genes  <- readLines("hugo_symbol_sex_genes.txt")

# Function to remove genes from RNA subsets that are found in the sex genes txt file
remove_sex_chromosome_genes <- function(rna, sex_chrom_genes) {
  
  xy_genes <- NULL
  # Check each gene in sex_chrom_genes
  for (gene in sex_chrom_genes) {
    
    # If gene found in RNA add to xy_gene list
    if (gene %in% rownames(rna)) {
      xy_genes <- append(xy_genes, gene)
    }
  }
  
  # Dimension of RNA before removing genes
  cat("Dimensions before removal:", dim(rna), "\n")
  
  # Remove genes
  rna <- rna[!(rownames(rna) %in% xy_genes), ]
  
  # Dimension of RNA after removing genes
  cat("Dimensions after removal:", dim(rna), "\n")
  
  # Return the updated RNA
  return((updated_rna = rna))
}

# Apply function to remove sex chromosome genes to both RNA subsets
rna <- remove_sex_chromosome_genes(rna, sex_chrom_genes)
```

# Plot the distribution of samples and genes before scaling and normalisation.
```{r}
library(ggplot2)
library(reshape2)

# Reshape the data to long format
rna_long <- melt(rna, variable.name = "Sample", value.name = "Expression")

# Create density plot with different lines for each sample
ggplot(rna_long, aes(x = Expression, color = Sample)) +
  geom_density() +
  labs(title = "Density Plots for All Samples", x = "Expression Values", y = "Density") +
  theme(legend.position = "none")

# Reshape the data to long format
rna_t <- data.frame(t(rna))
rownames(rna_t) <- colnames(rna)

# Subset to plot to only plot a subset of genes for an overview of their distributions
rna_t <- rna_t[,1:50]
rna_long_genes <- melt(rna_t, variable.name = "Gene", value.name = "Expression")

# Show distribution of genes
ggplot(data = rna_long_genes, aes(x = Expression, color = Gene)) +
  geom_density() +
  labs(title = "Density Plot for Genes", x = "Expression", y = "Density") +
  theme(legend.position = NULL)
```

# Subset data and normalise the subsets independantly to prevent dataleakage.
```{r}
# Load in IDs for each data subset
unsupervised_data <- readRDS("unsupervised_ID") #for MOFA
supervised_data <- readRDS("supervised_ID") #for single omic model
# Subset RNA data
rna_supervised <- rna[,colnames(rna)%in% supervised_data$X.Patient.ID ]
rna_unsupervised <- rna[,colnames(rna)%in% unsupervised_data$X.Patient.ID ]
```



```{r}
library(limma)

# log transform genes to remove the right skew of the data, making the data resemble more of a normal distribution
rna_supervised  <- log10(rna_supervised  + 1)
rna_unsupervised  <- log10(rna_unsupervised  + 1)
# Normalise in limma - using quantile normalise to make samples comparable
rna_supervised  <- data.frame(normalizeBetweenArrays(rna_supervised , method = "quantile"))
rna_unsupervised  <- data.frame(normalizeBetweenArrays(rna_unsupervised , method = "quantile"))

```

# Plot sample distribution
```{r}
# Reshape the data to long format
rna_long <- melt(rna_supervised, variable.name = "Sample", value.name = "Expression")
rna_long2 <- melt(rna_unsupervised, variable.name = "Sample", value.name = "Expression")

# Create density plot with different lines for each sample
ggplot(rna_long, aes(x = Expression, color = Sample)) +
  geom_density() +
  labs(title = "Density Plots for All Normalised Samples", x = "Expression Values", y = "Density") +
  theme(legend.position = "none")
ggplot(rna_long2, aes(x = Expression, color = Sample)) +
  geom_density() +
  labs(title = "Density Plots for All Normalised Samples", x = "Expression Values", y = "Density") +
  theme(legend.position = "none")

# Now the data is less right skewed 


# Plot gene distribution
# Reshape the data to long format
rna_t <- data.frame(t(rna_supervised))
rownames(rna_t) <- colnames(rna_supervised)
rna_t2 <- data.frame(t(rna_unsupervised))
rownames(rna_t2) <- colnames(rna_unsupervised)
# Subset to plot only a few genes
rna_t <- rna_t[,1:50]
rna_long <- melt(rna_t, variable.name = "Gene", value.name = "Expression")
rna_t2 <- rna_t2[,1:50]
rna_long2 <- melt(rna_t2, variable.name = "Gene", value.name = "Expression")

# Show distribution of genes
ggplot(data = rna_long, aes(x = Expression, color = Gene)) +
  geom_density() +
  labs(title = "Density Plot for Genes", x = "Expression", y = "Density") +
  theme(legend.position = NULL)
ggplot(data = rna_long2, aes(x = Expression, color = Gene)) +
  geom_density() +
  labs(title = "Density Plot for Genes", x = "Expression", y = "Density") +
  theme(legend.position = NULL)
```


# Extract the most varied features
```{r}
# Both subsets have to be considered at once, else different features will be extracted from each subset

# Function to extract most varied features
top_features <- function(rna1, rna2, nfeatures){
  if (nfeatures > nrow(rna)) {
  stop("nfeatures cannot be greater than the number of rows in the combined data.")
}
  # Combined into one df
  rna <- cbind(rna1,rna2)
  # Calculate the variance 
  var_rna <- apply(rna, 1, var, na.rm = TRUE)
  genes <- rownames(rna)
  variance_df <- data.frame(genes, var_rna)
  # Order the rows based on variance in decreasing order
  ordered_indices <- order(variance_df$var_rna, decreasing = TRUE)
  
  top_indices <- ordered_indices[1:nfeatures]
  top_varied_genes <- variance_df$genes[top_indices]
}

# Extract the top 5000 most varied genes considering both RNA subsets simultaneously
features <- top_features(rna_unsupervised,rna_supervised,5000)
length(unique(features))
# Subset both datasets to only include the top 5000 most varied genes
rna_unsupervised <- rna_unsupervised[features,]
rna_supervised <- rna_supervised[features,]

# Check for duplicates
sum(duplicated(rownames(rna_unsupervised)))
sum(duplicated(rownames(rna_supervised)))

# Save the preprocessed data
saveRDS(rna_unsupervised, "RNA_unsupervised_preprocessed")
saveRDS(rna_supervised, "RNA_supervised_preprocessed")

```

# Check the distribution of the most varied genes
```{r}
rna_t <- data.frame(t(rna_supervised))
rownames(rna_t) <- colnames(rna_supervised)
rna_t2 <- data.frame(t(rna_unsupervised))
rownames(rna_t2) <- colnames(rna_unsupervised)

# Subset to plot only a few genes
rna_t <- rna_t[,1:50]
rna_long <- melt(rna_t, variable.name = "Gene", value.name = "Expression")
rna_t2 <- rna_t2[,1:50]
rna_long2 <- melt(rna_t2, variable.name = "Gene", value.name = "Expression")

# Show distribution of genes
ggplot(data = rna_long, aes(x = Expression, color = Gene)) +
  geom_density() +
  labs(title = "Density Plot for Genes", x = "Expression", y = "Density") +
  theme(legend.position = NULL)
ggplot(data = rna_long2, aes(x = Expression, color = Gene)) +
  geom_density() +
  labs(title = "Density Plot for Genes", x = "Expression", y = "Density") +
  theme(legend.position = NULL)
```

##################################################################################
# Perform PCA - this is done on all samples so preprocessing is done on all samples together for this section

# Preprocess whole RNA dataset
```{r}
# Read in rna data
rna <- read.csv("ws3_grampian_rna_expression_median_per_gene.csv", row.names = 1)
# Remove the means column from the RNA dataset
rna <- rna[,-ncol(rna)]
# Read the txt file of gene names found on the x and y chromosomes
sex_chrom_genes  <- readLines("hugo_symbol_sex_genes.txt")

# Apply function to remove sex chromosome genes to both RNA subsets
rna <- remove_sex_chromosome_genes(rna, sex_chrom_genes)

library(limma)
# log transform genes to remove the right skew of the data, making the data resemble more of a normal distribution
rna  <- log10(rna  + 1)

# Normalise in limma - using quantile normalise to make samples comparable
rna  <- data.frame(normalizeBetweenArrays(rna , method = "quantile"))

# Extract the top 5000 most varied genes considering 
features <- top_features(rna,rna,5000)
length(unique(features))
rna <- rna[features,]
```

# Perform PCA
```{r}
library(ggplot2)
# Read in metadata
patient_data <- read.delim("ws3_grampian_patient_data.txt")

# Extract sample ID and Responses from metadata
metadata <- as.data.frame(patient_data[-c(1,2),c(1,20)])
dim(metadata)

# make sure samples overlap with the metadata
metadata <- metadata[metadata$X.Patient.ID %in% colnames(rna),]
dim(metadata)

# Perform pca
pca <- prcomp(t(rna), scale.=TRUE)
# Visualise pca
pca_data <- data.frame(Sample = colnames(rna), PC1 = pca$x[,1],PC2 = pca$x[,2], Response = metadata$Response.to.Treatment)


ggplot(pca_data, aes(x=PC1, y=PC2, color = factor(Response)))+
  geom_point(size = 1)+
  ggtitle("RNA Dataset")+
  stat_ellipse(geom = "polygon",
               aes(fill = Response), 
               alpha = 0.25)+
  scale_fill_brewer(palette = "Spectral")+
  scale_color_brewer(palette = "Spectral")+
  theme(plot.title = element_text(hjust = 0.5))+
  labs(color = "Response to Treatment")


```


